---
title: "Recreating the Effects of Hurricane Hugo on Stream Water Quality"
author: 
    name: Kailani Latimer
    email: kailani@ucsb.edu
date: "August 28, 2026"
toc: true
format: html
execute: 
    message: false
    warning: false
---

## Background
In 1989, Hurricane Hugo struck Puerto Rico. Scientists have known that stream water chemistry responds substantially to watershed disturbances. However, there's been no research on the effects of hurricanes on stream water quality. In the "Effects of hurricane disturbance" (Schaefer et al.) article, scientists look at the effects of Hurricane Hugo in several watersheds in Puerto Rico, Bisley Quebrada 1, 2, and 3, and Puente Roto Mameyes. Nitrate, potassium, and ammonium concentrations increased after the hurricane and remained elevated for a few years. Other ions concentrations' were elevated as well, but on a smaller scale. Overall, the findings concluded that hurricanes can have a greater effect on watersheds than other disturbances. 

## Data 
To investigate the datasets and to recreate Figure 3 from the article, I read in the data and cleaned the datasets so only the relevant data is leftover. The data included stream water chemistry from Quebrada one-Bisley (Q1), Quebrada two-Bisley (Q2), Quebrada three-Bisley (Q3), and Puente Roto Mameyes (MPR). Then, I cleaned that data, formatted it the correct way, and created a new file with a dataframe in it to reference later. 
```{r}
#| echo: false
#| include: false
library(tidyverse)
bisley_data <- read_csv("../output/cleaned_data.csv")
```


## Methods
First, I added the relevant datasets and created new clean dataframes for each dataset. The cleaned dataframes included the date and ions, dates from the beginning of 1988 to the end of 1994, and a renamed column called "Date". After adding the relevant datasets and cleaning them, I made a function to calculate the nine-week moving averages for all of the watersheds and their ion concentrations, so I didn't have to do each one individually. Then, in my cleaned file, I referenced the function for each of my cleaned dataframes. This created more new dataframes with the moving averages for each ion over the time period. Next, I combined these new dataframes, binding by rows, and pivoted it. Finally, after this binding and pivot, I created a new file that can be read in, without including all the code that went with it. This is helpful to make a plot like the one below.  


## Results
After calculating the moving averages, and joining the datasets together, I was able to recreate Figure 3 from the article, as shown in the plot below. On the x-axis is time in years with a vertical line representing when Hurricane Hugo hit. On the y-axis, is each individual ion's concentration with each separate graph representing it's respective ion. Within each ion's moving average, there are multiple lines with respective colors to each watershed. After Hurricane Hugo hit, ion concentrations spiked and remained elevated for years afterwards. This showcases that hurricanes are a large disturbance for stream water chemistry. 

```{r}
#| echo: false
#| warning: false
#| fig-height: 6
ggplot(
  data = bisley_data,
  mapping = aes(
    x = window_start,
    y = Concentration,
    color = site
  )
) +
  geom_line() +
  theme_bw() +
  labs(
    title = "Ion Concentration According to Watershed Around Hurricane Hugo",
    x = "Years",
    y = "Concentration",
    caption = "Concentrations in Bisley, Puerto Rico streams before and after Hurricane Hugo, 9-wk 
    moving averages. (a) calcium, (b) potassium, (c) magnesium, (d) ammonium-N and (e) nitrate-N. 
    The vertical lines mark the time of hurricane disturbance."
  ) +
  facet_wrap(
    ~Ions,
    scales = "free",
    ncol = 1,
    strip.position = "left",
    labeller = as_labeller(c(
      ca_mgl = "Ca mg l^-1",
      k_mgl = "K mg l^-1",
      mg_mgl = "Mg mg l^-1",
      nh4n_mgl = "NH_4-N ug l^-1",
      no3n_mgl = "NO_3-N ug l^-1"
    ))
  ) +
  geom_vline(xintercept = ymd("1989-09-18"), linetype = "dashed") +
  theme(
    plot.caption = element_text(hjust = 0),
    plot.caption.position = "plot"
  ) +
  scale_color_discrete("Watershed")
```